% !TEX root = chapter-standalone.tex


\chaptersecond{\statusdraft{Operads}}{chapter-operads}{\XXX}{
\todo{write}
}
\label{ch:operads} 



\begin{ctdefinition}[Operad]
  \label{def:operad}
  An \emph{\iindex{operad}} is defined by:
  
  \constit
  \begin{compactenum}
    \item \emph{Objects:} A collection~$\Ob_\operada$;
    \item \emph{Morphisms:} Let~$n\in \natnumbers$. For each finite string~$[\Obja_1,\ldots, \Obja_n]$ of objects and each object~$	\Objb$, one specifies a set~$\HomSet{\operada}{[\Obja_1,\ldots,\Obja_n]}{\Objb}$,
    elements of which are morphisms~$[\Obja_1,\ldots,\Obja_n]\mto \Objb$;
    \item \emph{Identity morphisms:} For each object~$\Obja$, a morphism~$\catid_\Obja \in \HomSet{\operada}{[\Obja_1,\ldots, 		\Obja_n]}{\Objb}$;
    \item \emph{Composition operations:}
    \begin{equation}
      \begin{aligned}
        \HomSet{\operada}{[\Obja_{1}^1, \ldots, \Obja_{n_1}^1]}{\Objb_1}\times \ldots \times  \HomSet{\operada}{[\Obja_{1}^m, \ldots, \Obja_{n_m}^m]}{\Objb_m}\times  \HomSet{\operada}{[\Objb_{1}, \ldots,\Objb_{m}]}{\Objc}&
        \mto  \HomSet{\operada}{[\Obja_{1}^1, \ldots, \Obja_{n_m}^m]}{\Objc}\\
        \tup{\mora_1,\ldots,\mora_m,\morb}&\mapsto [\mora_1,\ldots, \mora_m]\mthen \morb.
      \end{aligned}
  \end{equation}
  \end{compactenum}
  \condit
  \begin{compactenum}
    \item \emph{Associativity:}
    \begin{equation}
      [[\mora_{1}^1,\ldots, \mora_{n_1}^1]\mthen \morb_1,[\mora_{1}^2,\ldots, \mora_{n_2}^2]\mthen \morb_2,\ldots, [\mora_{1}^m,\ldots, \mora_{n_m}^m]\mthen \morb_m ]\mthen \morc=[\mora_{1}^1,\ldots, \mora_{n_m}^m]\mthen ([\morb_{1},\ldots, \morb_{m}]\mthen \morc).
    \end{equation}
    \item \emph{Unitality}:
    \begin{equation}
      [\catid_{\Obja_1},\ldots, \catid_{\Obja_n}]\mthen \mora = \mora = \mora \mthen \catid_{\Objb}, \quad \forall \mora\colon [\Obja_1,\ldots,\Obja_n]\mto \Objb.
    \end{equation}
  \end{compactenum}
\end{ctdefinition}

\begin{definition}
  \label{def:operad_from_monoidal}
  Let~$\tup{\CatC,\mtimescat,\monunit}$ be a strict monoidal category. Its associated operad~$\operada_\CatC$ has:
  \begin{compactenum}
    \item \emph{Objects}: $\Ob_{\operada_\CatC}=\Ob_\CatC$;
    \item \emph{Morphisms}: $\HomSet{\operada_\CatC}{[\Obja_1,\ldots,\Obja_n]}{\Objb}=\HomSet{\CatC}{\Obja_1\mtimescat\ldots \Obja_n}{\Objb}$;
    \item \emph{Identity morphism}: $\catid_\Obja \in \HomSet{\operada_\CatC}{[\Obja]}{\Obja}=\catid_\Obja \in \HomSet{\CatC}{\Obja}{\Obja}$;
    \item \emph{Composition of morphisms:}
  \end{compactenum}
\end{definition}

\begin{definition}
  \label{def:functors_operads}
  Let~$\operada,\operadb$ be operads. A \emph{functor} between operads~$\funa\colon \operada \fto \operadb$  is composed of:
  \begin{compactenum}
    \item A function~$\funaob\colon \Ob_\operada\to \Ob_\operadb$;
    \item A function~$\funamor\colon \HomSet{\operada}{[\Obja_1,\ldots,\Obja_n]}{\Objb}\to \HomSet{\operadb}{[\funaob(\Obja_1),\ldots, \funaob(\Obja_n)]}{\funaob(\Objb)}$.
  \end{compactenum}
  Constituents must be compatible with composition and identity morphisms.
\end{definition}

\begin{definition}
  \label{def:algebra_operad}
  Let~$\operada$ be an operad. An \emph{algebra} for~$\operada$ is a functor of operads~$\operada \fto \Set$.
\end{definition}

\begin{gradedexercise}[\exname{MonoidsAsAlgebras}]\label{ex:MonoidsAsAlgebras}
Let $\operada$ be the following operad. It has just a single object, which we call $\star$.  For each natural number $n \in \natnumbers$ (including $0$), we let $[n]$ denote the list consisting of $n$ copies of the object $\star$ (so $[0]$ in this case is a synonym for the empty list). With this notation, we define $\Hom_\operada([n];1)$ to be the 1-element set $\{ * \}$ for each $n \in \natnumbers$. The intuition is that, for each $n$, $\Hom_\operada([n];1)$ represents a single $n$-ary operation. 

In this exercise, the task is to describe what an algebra for this operad is! Try to prove as many of the statements that you make as you can. 
\end{gradedexercise}

\begin{gradedexercise}[\exname{MonoidActionsAsAlgebras}]\label{ex:MonoidsActionsAsAlgebras}
For the duration of this exercise, we fix a monoid $\monA$. 
We define an $\operada$ as follows. It has a single object, which we call $\star$.  For each natural number $n \in \natnumbers$ (including $0$), let $[n]$ denote the list consisting of $n$ copies of the object $\star$. We define $\Hom_\operada([n];1)$ to be the set underlying $\monA$ if $n =1$, and we set $\Hom_\operada([n];1) = \emptyset$ if $n \neq 1$. 

Can you describe what an algebra for this operad is?  Please prove your statements as best you can. 
\end{gradedexercise}






%\begin{definition}
%  \label{def:cospan}
%  Let~$\CatC$ be a category. A \emph{cospan} in~$\CatC$ is a pair of morphisms to a common object:
%  \begin{equation}
%    \includesag{cospan}
%  \end{equation}
%\end{definition}




\devel{
\begin{forslides}
  \begin{equation*}
\label{eq:composition_operad}
    [\mora,\morb]\mthen \morc
\end{equation*}
\end{forslides}

}

