% !TEX program = latexmk
\documentclass[paper=8.125in:10.250in,pagesize=pdftex,
    headinclude=false,footinclude=false,oneside,egregdoesnotlikesansseriftitles]{kaobook}

\input{utils/preamble}
\input{utils/body-init}

\title{ACT4E -- Draft of Exam 2}

\begin{document}
    \setboolean{bookversion}{true}
    \setboolean{statuscolors}{false}
    \setboolean{debugimages}{true}
    \setboolean{cachepdf}{true}
    \setboolean{devel}{true}
    \setboolean{showslides}{false}
    \setboolean{instructors}{true}
    \setboolean{codeexercises}{true}

\maketitle

%\section*{Meta}
%
%\noindent
%Exam date: 27.08.2021 \\
%Exam time: 10:00 -- 11:30  (1.5 hours) \\ 
%Exam place: HG E 7 \\
%Allowed materials: open book (laptops, phones, internet?)
%
%\
%
%\noindent
%Aims of the exam: 
%\begin{enumerate}
%\item Test familiarity with and understanding of basic definitions from the course
%\item Test proficiency in basic mathematical skills relating to proving statements, reasoning, and problem-solving
%\item Test the understanding of interconnections between various concepts introduced in the course, and the ability to transfer knowledge
%\item Test the understanding of the "compositional way of thinking" and the use of category theory concepts in connection with applied examples
%\end{enumerate}
%
%\
%
%\noindent
%Desired properties of the exam: 
%\begin{enumerate}
%\item Have more questions than are expected by any average or slightly above average student to be solved in the given 1.5 hours (we can then grade on a curve), in order to account for uncertainty in how difficult we judge the exam to be. 
%\item On the other hand we should not have too too many questions, because otherwise the grading work will be disproportionally large in proportion to the number of students. 
%\item Include some multiple choice questions as a way to provide students with a quick means to gain points and to reduce correction work. 
%\item Include some problems that ask for mathematical proofs. 
%\item Include some problems that have an applied flavor. 
%\item Include at least one problem that is challenging and requires some creative problem solving. 
%\end{enumerate}


\newpage

\section*{Exam information}


Welcome to the exam for the course Applied Compositional Thinking! 

\

\emph{Please do not look at the exam questions until instructed to begin the exam.}

\

Exam information: 
\begin{enumerate}
\item The exam will last 1.5 hours. We will tell you when you may begin, and we will signal when the time is up. 
\item Please write in blue or black pen. 
\item All pages with your answers on them must be labeled with your name!
\item Turn in all relevant sheets of paper at the end (including calculations, partial proofs, etc.) so that we consider as much of your work as possible when grading the exam. 
\item This is an ``open book'' exam. This means you can consult any books or other materials during the exam, including electronic PDFs and searching in the internet. However it is * NOT * allowed to communicate with anyone during the exam, electronically or otherwise. 
\end{enumerate}

\newpage


%\section*{Miscellaneous exam question ideas}
%
%\begin{exercise}
%Let $(\posA, \leq)$ be a poset. Explain how $(\posA, \leq)$ can be viewed as a category, and prove that, in this interpretation, it satisfies the definition of a category. 
%\end{exercise}
%
%
%\begin{exercise}
%Let $\tup{\monA, \then, \id}$ be a monoid. Explain how $\tup{\monA, \then, \id}$ can be viewed as a category, and prove that, in this interpretation, it satisfies the definition of a category. 
%\end{exercise}
%
%
%\begin{exercise}
%Something about monoid or group actions... maybe proving that the ``traditional definition'' is equivalent to one that involves the concept of ``functor'' ?
%\end{exercise}
%
%
%\begin{exercise}
%Check that the Kleisli category of a monad is indeed a category... (do this generically or using a concrete example?)
%\end{exercise}
%
%\begin{exercise}
%Check that the Eilenberg-Moore category of a monad is indeed a category...  (do this generically or using a concrete example?)
%\end{exercise}
%
%
%\begin{exercise}
%A problem about interpreting string diagrams for monoidal categories. 
%\end{exercise}
%
%
%\begin{exercise}
%More stuff using $Vect_R$ !
%\end{exercise}
%
%\newpage



\section*{Draft Exam 2}


\begin{gradedexercise}[\exname{LatticeCheck}]\label{ex:LatticeCheck}
(6 points) Here is a list of some posets. Which of them are lattices? 
Please give yes/no answers, and additionally, in the case of a "no", state all reasons why the poset in question fails to be a lattice. 
\begin{enumerate}
\item The poset given by the real numbers $\reals$ equipped with the usual ordering. 
\item The poset given by the set $\setA = \{ 1, 2, 3, 4, 6\}$ equipped with the ordering given by divisibility. That means, for all $\ela, \elb \in \setA$: $\ela \leq \elb$ if and only if $\ela$ divides $\elb.$ 
\item The poset given by the set $\{ \subA \subseteq \{1, 2 , 3 \} \mid 3 \notin \subA \}$, equipped with the ordering given by inclusion of subsets. 
\item Let $V$ be a finite-dimensional real vector space. Consider the poset given by the set $\mathcal{S}$ of linear subspaces of $V$, equipped with the ordering given by inclusion of subspaces. 
\end{enumerate}
\end{gradedexercise}

\solutionof{LatticeCheck}

\newpage

\begin{gradedexercise}[\exname{FreeCatOnGraph}]\label{ex:FreeCatOnGraph}
(5 points) Consider the following five graphs. In the course, we introduced the notion of the \emph{free category on a graph}. If $G$ is a graph, we denote the free category on $G$ by $\Free(G)$. For each graph $G$ below, how many morphisms in total are there in the associated category~$\Free(G)$?

  \begin{center}
    \includesag{20_dpcatfig_example_graphs}
  \end{center}
\end{gradedexercise}

\solutionof{FreeCatOnGraph}

\newpage


\begin{gradedexercise}[\exname{DirectSumProd}]\label{ex:DirectSumProd}
(9 points) Let $\Rel$ denote the category whose objects are sets and whose morphisms are binary relations. Given sets $\setA$ and $\setB$, we define their disjoint union to be the set
$$
\setA + \setB := \{ \tup{\ela, 1} \mid \ela \in \setA \} \cup \{ \tup{\elb, 2} \mid \elb \in \setB \}.
$$

Prove that in $\Rel$, the disjoint union -- together with projection morphisms which you should specify -- is ``the'' categorical product. 
\end{gradedexercise}

\solutionof{DirectSumProd}

\newpage


\begin{gradedexercise}[\exname{ProbFunc}]\label{ex:ProbFunc} (4 points).


Let $\Set$ denote the category of sets and functions. 
Given a set $\setA$, let $\mathcal{P} \setA$ denote the set of functions $p: \setA \to [0,1]$ which 
\begin{itemize}
\item take a non-zero value at only finitely-many elements of $\setA$, and 
\item satisfy the equation $\sum_{\ela \in \setA} p(\ela) = 1$. 
\end{itemize}
We think of $\mathcal{P} \setA$ as the set of finitely supported probability measures on $\setA$, or, in other words, the set of ``finite normalized histograms over the elements of $\setA$''. 

Given a function $\mapa: \setA \to \setB$ between sets, let $\mathcal{P} \mapa: \mathcal{P}\setA \to \mathcal{P}\setB$ be the function defined by 
$$ (\mathcal{P} \mapa)(p)(\elb) = \sum_{\ela \in \mora^{-1}(\{ \elb \})} p(\ela) \quad \quad  p \in \mathcal{P}\setA, \  \elb \in \setB.$$

Prove that the above defines a functor $\mathcal{P}: \Set \to \Set$. 

Remark: you do not need to prove that, given $p \in \mathcal{P}\setA$, the function $(\mathcal{P} \mapa)(p)$ is indeed an element of $\mathcal{P}\setB$; you can assume this fact to be true. 
\end{gradedexercise}

\solutionof{ProbFunc}

\newpage


\begin{gradedexercise}[\exname{ProbNatTrafos}]\label{ex:ProbNatTrafos} (5 points)


Let $\Set$ be the category of sets, let $\mathbf{Id}_\Set : \Set \to \Set$ be the identity functor, and let $\mathcal{P}: \Set \to \Set$ be the functor defined in \cref{ex:ProbFunc}. 
\begin{enumerate}
\item Given a set $\setA$, let $\delta_{\setA} : \setA \to \mathcal{P} \setA$ be the function defined by
$$\delta_\setA (\ela)(\elb) = \left\{\begin{array}{ll}
        1 & \text{if } \elb = \ela \\
        0 & \text{else.} 
        \end{array}\right. 
  \quad \quad \ela, \elb \in \setA.$$
Prove that the collection of functions $\{ \delta_{\setA} \}_{\setA \in \Ob_{\Set}}$ defines a natural transformation $\delta: \mathbf{Id}_\Set \Rightarrow \mathcal{P}$. 
\item Given a set $\setA$, let $\mu_\setA : \mathcal{P} \mathcal{P} \setA \to \mathcal{P} \setA$ be the function defined by 
$$\mu_\setA (f)(\ela) = \sum_{p \in \mathcal{P} \setA} f(p) p(\ela) \quad \quad f \in \mathcal{P} \mathcal{P} \setA, \ela \in \setA.$$
Prove that the collection of functions $\{ \mu_{\setA} \}_{\setA \in \Ob_{\Set}}$ defines a natural transformation $\mu: \mathcal{P} \mathcal{P} \Rightarrow \mathcal{P}$. 
\end{enumerate}
\end{gradedexercise}

\solutionof{ProbNatTrafos}

\newpage


\begin{gradedexercise}[\exname{ProbMonad}]\label{ex:ProbMonad} (7 points).


Let $\Set$ denote the category of sets, and let $\mathcal{P}$, $\mathbf{Id}_\Set$, $\delta$, and $\mu$ be as in \cref{ex:ProbFunc} and \cref{ex:ProbNatTrafos}. 
Prove that $\tup{\mathcal{P}, \mu, \delta}$ defines a monad on $\Set$. 
\end{gradedexercise}

\solutionof{ProbMonad}

\newpage


\end{document}
