% !TEX program = latexmk
\documentclass[paper=8.125in:10.250in,pagesize=pdftex,
    headinclude=false,footinclude=false,oneside,egregdoesnotlikesansseriftitles]{kaobook}

\input{utils/preamble}
\input{utils/body-init}

\title{ACT4E Exam Questions Pool}

\begin{document}
    \setboolean{bookversion}{true}
    \setboolean{statuscolors}{false}
    \setboolean{debugimages}{true}
    \setboolean{cachepdf}{true}
    \setboolean{devel}{true}
    \setboolean{showslides}{false}
    \setboolean{instructors}{true}
    \setboolean{codeexercises}{true}

\maketitle

%\section*{Meta}
%
%\noindent
%Exam date: 27.08.2021 \\
%Exam time: 10:00 -- 11:30  (1.5 hours) \\ 
%Exam place: HG E 7 \\
%Allowed materials: open book (laptops, phones, internet?)
%
%\
%
%\noindent
%Aims of the exam: 
%\begin{enumerate}
%\item Test familiarity with and understanding of basic definitions from the course
%\item Test proficiency in basic mathematical skills relating to proving statements, reasoning, and problem-solving
%\item Test the understanding of interconnections between various concepts introduced in the course, and the ability to transfer knowledge
%\item Test the understanding of the "compositional way of thinking" and the use of category theory concepts in connection with applied examples
%\end{enumerate}
%
%\
%
%\noindent
%Desired properties of the exam: 
%\begin{enumerate}
%\item Have more questions than are expected by any average or slightly above average student to be solved in the given 1.5 hours (we can then grade on a curve), in order to account for uncertainty in how difficult we judge the exam to be. 
%\item On the other hand we should not have too too many questions, because otherwise the grading work will be disproportionally large in proportion to the number of students. 
%\item Include some multiple choice questions as a way to provide students with a quick means to gain points and to reduce correction work. 
%\item Include some problems that ask for mathematical proofs. 
%\item Include some problems that have an applied flavor. 
%\item Include at least one problem that is challenging and requires some creative problem solving. 
%\end{enumerate}
%

\newpage

\section*{Miscellaneous exam question ideas}

\begin{exercise}
Let $(\posA, \leq)$ be a poset. Explain how $(\posA, \leq)$ can be viewed as a category, and prove that, in this interpretation, it satisfies the definition of a category. 
\end{exercise}


\begin{exercise}
Let $\tup{\monA, \then, \id}$ be a monoid. Explain how $\tup{\monA, \then, \id}$ can be viewed as a category, and prove that, in this interpretation, it satisfies the definition of a category. 
\end{exercise}


\begin{exercise}
Something about monoid or group actions... maybe proving that the ``traditional definition'' is equivalent to one that involves the concept of ``functor'' ?
\end{exercise}


\begin{exercise}
Check that the Kleisli category of a monad is indeed a category... (do this generically or using a concrete example?)
\end{exercise}

\begin{exercise}
Check that the Eilenberg-Moore category of a monad is indeed a category...  (do this generically or using a concrete example?)
\end{exercise}


\begin{exercise}
A problem about interpreting string diagrams for monoidal categories. 
\end{exercise}


\begin{exercise}
More stuff using $Vect_R$ !
\end{exercise}

\newpage


\section*{Exam 	Questions Pool}




\begin{gradedexercise}[\exname{RelKleisli}]\label{ex:RelKleisli} (13 points) 

Let $\Set$ denote the category of sets and $\powerset : \Set \to \Set$ the powerset functor which assigns, to any given set $\setA$, the set of subsets of $\setA$. The endofunctor $\powerset$ may be equipped with the structure of a monad $(\powerset, \eta, \mu)$, where the components of the natural transformations $\eta$ and $\mu$ are given, respectively, by the functions
$$ \eta_{\setA} : \setA \to \powerset \setA, \ \ela \mapsto \{ \ela \}$$
and
$$ \mu_{\setA} : \powerset \powerset \setA \to \powerset \setA, \ \sigma \mapsto \bigcup_{S \in \sigma} S.$$

\begin{enumerate}
\item (2 points) Prove that $\eta$, as defined above, is in fact a natural transformation. 
\item Let $\Set_{\powerset}$ denote the Kleisli category of the monad $(\powerset, \eta, \mu)$. The aim of this part of the exercise is to show that the category $\Set_{\powerset}$ and the category $\Rel$ of sets and relations are isomorphic as categories. Your tasks are:
\begin{enumerate}
\item (8 points) Define functors $\funa: \Set_{\powerset} \to \Rel$ and $\funb: \Rel \to \Set_{\powerset}$ and prove that they are in fact functors. 
\item (3 points) Prove that $\funa$ is an isomorphism of categories, with inverse $\funb$. 
\end{enumerate}
\end{enumerate}
\end{gradedexercise}

\solutionof{RelKleisli}

\newpage





\end{document}
