\documentclass{book}
\usepackage{amsthm}
\usepackage{xparse}
\usepackage{environ}
\usepackage[colorlinks]{hyperref}

\newtheoremstyle{exercise}% name
  {\topsep}%     Space above
  {\topsep}%     Space below
  {\normalfont}% Body font
  {}%            Indent amount (empty = no indent, \parindent = para indent)
  {\bfseries}%   Thm head font
  {.}%           Punctuation after thm head
  { }%           Space after thm head
  {\hyperref[dvvv_#2_reverse_label]{\thmname{#1}\thmnumber{ #2}}\thmnote{ (#3)}}% Thm head spec

\theoremstyle{exercise} % body text is upright
\newtheorem{innerexercise}{Exercise}
\theoremstyle{definition}

\ExplSyntaxOn
\NewDocumentEnvironment{exercise}{}
 {
  \innerexercise\label{dvvv_ \theinnerexercise _label}
 }
 {
  \endinnerexercise
 }
\newtheorem*{innersolution}{\dvvv_exercise_ref:}

\NewEnviron{solution}
 {
  \seq_gput_right:Nx \g_dvvv_solution_group_seq
   {
    {\theinnerexercise}{\exp_not:V \BODY}
   }
 }
\NewDocumentEnvironment{solution*}{}
 {
  \phantomsection\label{dvvv_ \theinnerexercise _reverse_label}
  \cs_set:Npx \dvvv_exercise_ref:
   {
    \exp_not:N \hyperref
     [
      dvvv_ \theinnerexercise _label
     ]
     {
      Solution~of~\exp_not:N\ref*{dvvv_ \theinnerexercise _label}
     }
   }
  \innersolution
 }
 {
  \endinnersolution
 }
\NewDocumentCommand{\printsolutions}{}
 {
  \seq_map_inline:Nn \g_dvvv_solution_group_seq
   {
    \dvvv_print_solutions:nn ##1
   }
  \seq_gclear:N \g_dvvv_solution_group_seq
 }

\seq_new:N \g_dvvv_solution_group_seq

\cs_new_protected:Npn \dvvv_print_solutions:nn #1 #2
 {
  \cs_set:Npn \dvvv_exercise_ref:
   {
    \phantomsection\label{dvvv_#1_reverse_label}
    \hyperref[dvvv_#1_label]{Solution~of~\ref*{dvvv_#1_label}}
   }
  \begin{innersolution} #2 \end{innersolution}
 }
\ExplSyntaxOff

\begin{document}

\chapter{Title}

\section{Exercises}

\begin{exercise}
This is an exercise.
\end{exercise}
\begin{solution}
A solution.
\end{solution}

\begin{exercise}
This is an exercise.
\end{exercise}
\begin{solution}
A solution.
\end{solution}

\begin{exercise}
This is an exercise.
\end{exercise}

Some optional text in between.

\begin{solution*}
This is a solution.
\end{solution*}

The solution has been printed immediately, because
\texttt{solution*} has been used.

\begin{exercise}
This is another exercise.
\end{exercise}
\begin{solution}
A solution.
\end{solution}

\begin{exercise}
This is yet another exercise.
\end{exercise}
\begin{solution}
Too much.
\end{solution}

\clearpage
\section{Solutions}
\printsolutions

\end{document}
