% !TEX root = ../ACT4E-jonathan.tex
 
\section{Functor composition}

\JL{I think this should be a definition, not a lemma. }

In the following, we want to show that functors compose.
Given categories~$\Cat{A},\Cat{B},\Cat{C}$ and functors~$F\colon \Cat{A}\to \Cat{B}$,~$G\colon \Cat{B}\to \Cat{C}$, we want to show that~$F\then G$ is a functor. To do this, we show that~$F\then G$ preserves identities and compositions. 
\begin{itemize}
    \item Given an object~$\obja\in \Cat{A}$, we have:
    \begin{equation*}
        \begin{aligned}
            (F\then G)(\id_\obja)&=G(F(\id_\obja))\\
            &=G(\id_{F(\obja)})\\
            &=\id_{G(F(\obja))},
        \end{aligned}
    \end{equation*}
    where we used that~$F$ and~$G$ are functors (i.e., they preserve identities).
    \item Furthermore, given composable morphisms $f,g\in \Cat{A}$, one has:
    \begin{equation*}
        \begin{aligned}
            (F\then G)(f\then g)&=G(F(f)\then F(g))\\
            &=G(F(f))\then G(F(g)),
        \end{aligned}
    \end{equation*}
    where again used that~$F,G$ are functors (i.e., they preserve composition).
\end{itemize}


We can define an identity functor. Given a category~$\CatC$, we define it as~$\id_\CatC\colon \CatC \to \CatC$,~$\id_\CatC(x)=x$ for every object and morphism in~$\CatC$. To show that this is a valid functor, we need to show that it preserves identities and composition:
\begin{itemize}
    \item Given any~$\obja \in \Ob_\CatC$, we have:
    \begin{equation*}
        \begin{aligned}
            \id_\CatC(\id_\obja)&=\id_\obja\\
            &=\id_{\id_\CatC(\obja)}
        \end{aligned}
    \end{equation*}
    Furthermore, given composable morphisms~$f,g$ in~$\CatC$, we have:
    \begin{equation*}
    \begin{aligned}
        \id_{\CatC}(f\then g)&=f\then g\\
        &=\id_{\CatC}(f)\then \id_{\CatC} (g).
    \end{aligned}
    \end{equation*}
\end{itemize}

\section{A category of categories}

\todo{Hence we can define this category.}

\begin{ctdefinition}
  There is a category, called \Category, which is constituted of
  \begin{itemize}
    \item Objects: categories;
    \item Morphisms: functors;
    \item Identity morphisms: identity functors;
    \item Composition: composition of functors.
  \end{itemize}
\end{ctdefinition}
