% !TEX root = ../ACT4E-full.tex

We have seen that functors are ``morphisms between categories''. Indeed, categories may be assembled into a category \index{\Category} where the objects are categories and morphisms are functors. It turns out that there is an important third layer to this world of categories: there are also kinds of morphisms \emph{between} functors, and these are known as ``natural transformations''. To represent the three layers of structure involved in the world of categories, we will often draw diagrams like this,
\begin{equation*}
  \includesag{3-layer-diagrams}
\end{equation*}


where the points represent categories, the single arrows represent functors, and the double arrows represent natural transformations.


\begin{ctdefinition}[Natural transformation]
  Let \CatC and \CatD be categories, and let~$F,G\colon \CatC\to \CatD$ be functors. A \emph{\iindex{natural transformation}}~$\alpha\colon F\to G$
  \begin{equation}
    \includesag{55_natural_1}
  \end{equation}
  is defined by the following constituent data, satisfying the following condition.
  \underline{Data:}
  \begin{compactenum}
    \item For each object~$\obja\in \CatC$, a morphism $\alpha_\obja \colon F(\obja)\to G(\obja)$ in \CatD, called the $\obja$\emph{-component} of $\alpha$.
  \end{compactenum}
  \underline{Condition:}
  \begin{compactenum}
    \item For every morphism~$f\colon \obja\to \objb$ in \CatC, the components of $\alpha$ must satisfy the \emph{naturality condition}
    \begin{equation}
      F(f)\then \alpha_\objb = \alpha_\obja\then G(f),
    \end{equation}
    i.e. the following diagram must commute:
    \begin{equation}
      \includesag{55_natural_2}
    \end{equation}
  \end{compactenum}
\end{ctdefinition}

\begin{ctdefinition}[Natural isomorphism]
  \label{def:nat_iso}
  A natural transformation~$\alpha\colon F\to G$ is called a \emph{\iindex{natural isomorphism}} if each component~$\alpha_\obja$ is an isomorphism in \CatD.
\end{ctdefinition}

\begin{example}
  Consider the category~$\Vect_{\reals}$ whose objects are real vector spaces and whose morphisms are linear maps. (For convenience, in the following we sometimes omit reference to the ground field.) Recall that the \emph{dual} of a vector space~$V$ is the vector space
  \begin{equation*}
    V^* \coloneqq \Hom_{\Vect}(V, \reals),
  \end{equation*}
  i.e., the space of all linear maps from~$V$ to~$\reals$. Also, recall the if~$f\colon V \to W$ is a linear map, then its dual is a linear map~$f^*\colon W^* \to V^*$.

  Applying the above duality construction twice to a vector space or a linear map gives their double dual. It turns out that this is a functorial operation. That is, there is a functor~$\text{Double dual}\colon \Vect \to \Vect$ that maps every vector space and every linear map to its double dual.

  Furthermore, for any vector space~$V$, there exists a ``canonical'' or ``natural'' map~$\alpha_V \colon V \to V^{**}$ defined by
  \begin{equation*}
    \alpha_V(v)(l) = l(v), \quad  v \in V, l \in V^*.
  \end{equation*}
  These form the components of a natural transformation from the identity functor on \Vect to the double dual functor.
  \begin{equation}
    \includesag{nat-trafo-ddual}
  \end{equation}
\end{example}



