% !TEX root = chapter-standalone.tex

\section{What is ``design''?}
\label{sec:design-what-is-design}

%\linkvideo{spring2021-design:design} % Design
\linkvideo{spring2021-design:design:def-design} % What is design?

We take a broad view of what it means to ``design'', that is not limited to engineering.
Citing at length Hebert Simon's\footnote{Hebert A.
    Simon (1916-2001).
    Winner of the 1978 Nobel Prize in Economics.
}
\emph{The sciences of the artificial}~(\cite{hebert96sciences}, Chapter 5):

\begin{quote}
    Engineers are not the only professional designers.
    Everyone designs who devises courses of action aimed at changing existing situations into preferred ones.
    The intellectual activity that produces material artifacts is no different fundamentally from the one that prescribes remedies for a sick patient or the one that devises a new sales plan for a company or a social welfare policy for a state.
    Design, so construed, is the core of all professional training;
    it is the principal mark that distinguishes the professions from the sciences.
    Schools of engineering, as well as schools of architecture, business, education, law, and medicine, are all centrally concerned with the process of design.
\end{quote}

The metaphors used in the book are biased towards engineering.
It is easy for everybody to imagine creating a physical machine out of simple components, and what choices and trade-offs we must deal with.
Furthermore, it is easy to imagine what is the boundary between the machine and the world, that is, to delimit the design space.

Yet the theory to be discussed is applicable to other disciplines, if one takes a more abstract view of what is a system and a component.
For example, in urban planning, the components of a city are roads, sewers, residential areas, \etc
In other disciplines, ``components'' can be logical instead of physical.
For example, an economist might ask how to design an incentive scheme such that such scheme (a ``component'') will move the system to a more desirable set of states.
