 %% IF POCKET BOOK UNCOMMENT
%\documentclass[b5paper, onecolumn, superscriptaddress,shorttitle=papers]{compositionalityarticle}


\documentclass[a4paper,onecolumn, superscriptaddress,10pt,shorttitle=papers]{compositionalityarticle}
\pdfoutput=1
\usepackage[backend=bibtex]{biblatex}
\addbibresource{dpcatbib.bib}

% Including extra packages
\input{utils/packages.tex}
% Including symbols
\input{utils/symbols.tex}
%\tikzexternalize{prefix=sag/}% works locally, not on cloud
% Works if pdflatex -synctex=1 -interaction=nonstopmode --shell-escape %.tex

%\includeonly{chapters/10_introduction}
%\includeonly{chapters/15_basic_concepts}
%\includeonly{chapters/20_things_transforming,chapters/30_things_connecting,}
%\includeonly{chapters/15_basic_concepts,chapters/20_things_transforming,chapters/30_things_connecting,chapters/40_specificity_constraints,chapters/50_resource_composition,chapters/60_alternatives}
%\includeonly{chapters/70_tradeoffs,chapters/75_posets_with_more,chapters/80_monotonicity,chapters/95_functors}
%\includeonly{chapters/40_specificity_constraints}
%\includeonly{chapters/50_resource_composition}
%\includeonly{chapters/60_alternatives}
%\includeonly{chapters/70_tradeoffs}
%\includeonly{chapters/75_posets_with_more}
%\includeonly{chapters/80_monotonicity}
%\includeonly{chapters/95_functors}
%\includeonly{chapters/90_duality_design}
%\includeonly{chapters/100_composing_no_structure}
%\includeonly{chapters/110_composing_structure}
%\includeonly{chapters/150_restrictions}
%\includeonly{chapters/115_parallel_theory}
\includeonly{chapters/120_parallel_composition}
%\includeonly{chapters/130_design_problems}
%\includeonly{chapters/135_companion_and_conjoint}
%\includeonly{chapters/143_categorical_trace}
%\includeonly{chapters/146_closing_loop_codesign}
%\includeonly{chapters/160_ordering_dps}
%\includeonly{chapters/165_order_and_compositions}
%\includeonly{chapters/155_relations_facts}
%\includeonly{chapters/170_higher_order_design}
%\includeonly{chapters/180_uncertainty}
%\includeonly{chapters/190_enrichments}
%\includeonly{chapters/200_computations}


\begin{document}


\title{Mathematical Foundations of Engineering Co-Design}
\date{}
\author{Andrea Censi}
\email{acensi@idsc.mavt.ethz.ch}
\homepage{https://censi.science}
\orcid{TODO}
% \thanks{You can use the \texttt{\textbackslash{}email}, \texttt{\textbackslash{}homepage}, and \texttt{\textbackslash{}thanks} commands to add additional information for the preceding \texttt{\textbackslash{}author}. If applicable, this can also be used to indicate that a work has previously been published in conference proceedings.}
\author{Jonathan Lorand}
\email{jlorand@ethz.ch}
\homepage{http://lorand.earth/math}
\affiliation{Institute of Dynamic Systems and Control (IDSC),
    Department of Mechanical and Process Engineering (D-MAVT), ETH Z\"urich, Z\"urich,
    Switzerland}
\author{David I. Spivak}
\email{dspivak@gmail.com}
\homepage{http://math.mit.edu/$\sim$dspivak/}
\affiliation{Department of Mathematics, Massachusetts Institute of Technology,
    Cambridge, MA, USA.}
\author{Joshua Tan}
\email{joshua.tan@cs.ox.ac.uk}
\homepage{http://joshuatan.com/research}
\affiliation{Department of Computer Science, University of Oxford, Oxford, UK.}
\author{Gioele Zardini}
\email{gzardini@ethz.ch}
\homepage{http://gioele.science}
\maketitle

\begin{abstract}
    This paper is about engineering design and
    category theory: an odd couple, consisting of one of the most concrete and
    one of the most abstract fields of human knowledge.
    We will show how category theory can help in defining, understanding, and
    solving formal engineering design problems.
    This theory is especially well-suited for
    heterogeneous components with complex
   recursive co-design constraints.


    This paper is written for two audiences: we
    want to explain to engineers the power and the necessity of the
    category-theoretical language to describe complex design problems; and we
    want to call the mathematicians' attention to the correspondence between
    very abstract category theory and extremely concrete engineering design
    choices that are observable in any artefact.


    For the reader who already knows category theory: We introduce a category~$\DP$ of
    \emph{design problems} and study its properties. We show that~$\DP$ is a
    generalization of the category~$\textbf{Rel}$ of relations.  Diagrams
    in~$\DP$ describe the co-design constraints among different subsystems of a
    large system. We show that~$\DP$ is \emph{compact closed}, which makes the
    diagrams easy to manipulate as they behave similarly to other formalisms
    such as signal flow diagrams. Finally, we describe the 2-category structure
    of~$\DP$, which models naturally the dominance between alternative designs.
\end{abstract}

\clearpage

\tableofcontents
% Introduction
\include{chapters/10_introduction}
\clearpage
% Basic concepts of formal engineering design
\include{chapters/15_basic_concepts}
\clearpage
% How things transform
\include{chapters/20_things_transforming}
\clearpage
% How things connect
\include{chapters/30_things_connecting}
% Constraints
\include{chapters/40_specificity_constraints}
\clearpage
% Resource composition
\include{chapters/50_resource_composition}
\clearpage
% Alternatives
\include{chapters/60_alternatives}
\clearpage
% Tradeoffs
\include{chapters/70_tradeoffs}
\clearpage
% Posets with more structire
\include{chapters/75_posets_with_more}
\clearpage
% Monotonicity
\include{chapters/80_monotonicity}
\clearpage
% Functors (TBD)
\include{chapters/95_functors}
\clearpage
% Duality of design
\include{chapters/90_duality_design}
\clearpage
% Composing without structure
\include{chapters/100_composing_no_structure}
\clearpage
% Composing recognizing structure
\include{chapters/110_composing_structure}
\clearpage
% Restrictions and alternatives
\include{chapters/150_restrictions}
\clearpage
% Parallel composition theory
\include{chapters/115_parallel_theory}
\clearpage
% Parallel composition
\include{chapters/120_parallel_composition}
\clearpage
% Defining design problems
\include{chapters/130_design_problems}
\clearpage
% Companion and Conjoint
\include{chapters/135_companion_and_conjoint}
\clearpage
% Cartegorical Trace
\include{chapters/143_categorical_trace}
\clearpage
% Closing the loop in co-design problems
\include{chapters/146_closing_loop_codesign}
\clearpage
% Ordering design problems
\include{chapters/160_ordering_dps}
\clearpage
% Orders and composition
\include{chapters/165_order_and_compositions}
\clearpage
% Relations, facts
\include{chapters/155_relations_facts}
\clearpage
% Higher order design problems
\include{chapters/170_higher_order_design}
\clearpage
% Uncertainty
\include{chapters/180_uncertainty}
\clearpage
% Enrichment
\include{chapters/190_enrichments}
\clearpage
% Computability
\include{chapters/200_computations}
\clearpage

% --- OLD MATERIAL TO PUT BACK UP
%Introduction
 \include{chapters/20_introduction}
\clearpage
% Background in CT
\include{chapters/30_basic_category_theory}
\clearpage
% DP Traced monoidal
\include{chapters/50_DP_traced_monoidal}
\clearpage
% Creating DPs
\include{chapters/60_creating_DPs}
\clearpage

\printbibliography

\end{document}
