\section{Minimal and maximal elements}

You know already the operators~$\min$/$\max$ that give the minimum/maximum values of a set of real numbers.
If the set is finite, the minimum and maximum always exist.
But for infinite sets, the minimum and maximum might not exist.
For example, consider the set of real numbers contained between 0 and 1, excluding the boundaries:
%
\begin{equation}
    \setA = \makeset{ x \setin \reals \colon 0 < x < 1 }.
\end{equation}
%
This set does not have a minimum or a maximum.

For a total order, if the minimum and maximum exist, then they are unique.
In a partial order, this is not the case.
We introduce the operators~$\Min$ and~$\Max$ that are the generalization to partial orders of~$\min/\max$.

\begin{ctdefinition}[Minimal elements]
    \label{def:Min}
    $\Min \colon \posPA \to \posAA$ is the map that sends a subset~$\subA$ of a \SY{poset} to the minimal elements of that subset (those elements~$\setAel \setin S$ such that~$\setAel \posAleq \setBel$ for all~$\setBel \setin \subA$).
    In formulas:
    \begin{equation}
        \defmapperiodset{\Min}{\posPA}{\posAA}{\subA}{\makesett{
                \setCel\setin \subA\colon
                \vmiddle{
                    \prftree{
                        \setDel\setin \subA
                    }{
                        \setDel\posAleq \setCel
                    }{
                        \setCel=\setDel
                    }
                }
            }}
    \end{equation}
    Note that~$\Min(\subA)$ could be empty.
\end{ctdefinition}

\begin{ctdefinition}[Maximal elements]
    \label{def:Max}
    $\Max \colon \posPA \to \posAA$ is the map that sends a subset~$\subA$ of a \SY{poset} to the maximal elements of that subset (those elements~$\setAel \setin \subA$ such that~$\setAel \posgeqof\posA \setBel$ for all~$\setBel \setin \subA$).
    In formulas:
    \begin{equation}
        \defmapperiodset{
            \Max
        }{
            \posPA
        }{
            \posAA
        }{
            \subA
        }{
            \makesett{
                \setCel\setin \subA\colon
                \vmiddle{
                    \prftree{
                        \setDel\setin \subA
                    }{
                        \setDel\posAgeq \setCel
                    }{
                        \setCel=\setDel
                    }
                }
            }
        }
    \end{equation}
    Note that~$\Max(\subA)$ could be empty.
\end{ctdefinition}

