% !TEX root = chapter-standalone.tex
\section{Set up}

\subsection{Sign up on Github}

Sign up on Github if you haven't already.

\subsection{Option 1: Using Github Classroom}\label{subsec:with-classroom}

\begin{quote}
  \emph{
    If you are following a class where the instructors are using Github Classroom,
    this is the section for you. Otherwise, jump to \cref{sub:without-classroom}.}
\end{quote}

Github Classroom is a product that helps coordinate assignments for large classes.

Your instructors will send you an invite link to join the classroom. An invite link is similar to this:

\url{https://classroom.github.com/a/XXXXXXX}

If you are not enrolled in an official class, then you can get the invite link on Zulip.

In any case, do not pass invite links around: there are different ``classrooms''.
Do not use old links because the classroom is refreshed after each season.

Once you go through the procedure, you will have created a repository

\begin{quote}
\str{https://github.com/INSTRUCTORS/solution-YOU}
\end{quote}
where \str{INSTRUCTORS} is the Github organization for the instructors, and \str{YOU} is your username.

Clone the repository just created and jump into the directory.

You should see the following content:
%
%% Trick so that pycharm does not reformat inside
%\begin{minted}[autogobble]{console}
% .  src/
% .   act4e_solutions/
% .      __init__.py
% .  Dockerfile
% .  Makefile
% .  requirements.txt
% .  setup.py
%\end{minted}
%
%\dirtree{
%[root
%  [src
%    [act4e\_solutions.py
%    [\_\_init\_\_.py]
%    ]
%  ]
%  [Dockerfile]
%  [Makefile]
%  [requirements.txt]
%  [setup.py]
%]
%}
\includesag{dirtree-setup}

\subsection{Option 2: Independent use without Github Classroom}
\label{sub:without-classroom}

\begin{quote}
  \emph{
    If you are \emph{not} following a class where the instructors are using Github Classroom,
    this is the section for you. Otherwise, jump to \cref{subsec:with-classroom}.
  }
\end{quote}

In this section we assume that you are not part of a class.

In that case, you can fork directly the repository

\url{https://github.com/ACT4E/ACT4E-exercises-template}

Note that there might be different branches to choose from. The instructions in this book correspond to
the branch \texttt{spring2021}.

You can do the exercises on your own by using this repository.

In the following, ignore any other mention of Github Classroom.

\subsection{Docker installation}

Install Docker following the information at

\url{https://docs.docker.com/get-docker/}

